Learning with Misspecified Models: the case of overestimation

### 2023-10-14 Paper/draft.Rmd
Edit intro and abstract

---
title: "Learning with misspecified models: the case of overestimation"  
date: '`r format(Sys.time(), "%B %d, %Y")`'
author: "Jimena Galindo"
abstract: "I design a framework and laboratory experiment that allows for the comparison of multiple theories of misspecified learning. I focus on a framework with endogenous information and two a data generating process ruled by two fundamentals: an ego-relevant parameter and a state. Within this framework I study three forces that can lead to misspecified beliefs: initial misspecifications, learning traps and biased updating. I find that biased updating is the main driver of misspecifications. In addition, I vary the degree of ego-relevance of the parameter by introducing a stereotype treatment. The data is consistent with biased updating in both cases but for potentially different reasons: In the ego-relevant treatment subjects attribute successes to their own ability and failures to luck; in the stereotype treatment they over-compensate for initial negative biases by over-attributing positive signals to the ability of others. This translates into similar observed choices but different dynamic in beliefs."
#classoption: pagebackref # passes pagebackref=true to hyperref before it is loaded, allowing backreferencing
output: 
    pdf_document:
        citation_package: natbib
        fig_width: 7  
        fig_height: 6
        fig_caption: true
        number_sections: true

@@

A growing body of literature in economics explores how people develop incorrect beliefs 
about fundamentals. Most of this research centers on scenarios where agents 
passively observe the world, and incorrectly integrate the received information into their 
beliefs.\footnote{See \citet{benjamin2019} for a review of the literature on errors of 
probabilistic thinking}
However, many real-world situations cast agents as active participants in the generation of 
information. In these cases, the information they acquire is influenced by their actions and 
subsequent behavior is in turn determined by how they incorporate the information 
into their beliefs.

Consider a student who needs to decide how much effort to put into 
studying for an exam. 
Their decision will depend on two factors: their belief about their intrinsic 
ability, and their belief about how difficult the exam will be. The outcome 
they observe will be affected by how much they decide to study. Imagine that the student 
puts in a moderate amout of effort and gets a surprisingly good grade. Did they get a good grade because 
they are smarter than they thought? Or because the exam was easier than they had anticipated? 
Their future exam preparation strategy will depend fundamentally on which line of reasoning they 
take. This feedback loop is refered to as an \emph{endogenous information 
process}.

I compare a set of theories that model learning in an endogenous information setting and 
which can rationalize 
the persistence of misspecified beliefs. I describe a unifying framework that nests multiple 
theories of model learning and generates testable predictions for each of them. 
I cast this framework in a laboratory experiment design and test these predictions to identify which 
of these theories are consistent with the behavior I observe in the lab.
The framework features an agent who needs to 
learn two parameters: one that regards their own characteristics 
(an \emph{ego-relevant} parameter), 
and an exogenously determined state. In the context of the student preparing for exams, 
the two parameters  would be their intrinsic ability and the difficulty of the exam.

@@

In all of these examples,
holding an incorrect belief about a fundamental leads to suboptimal choices with 
potentially high costs. In spite of the abundance of evidence, the scope of the existing research 
is limited in terms of the frameworks it considers. Most of the experimental evidence 
documenting the bias is collected in settings where subject are passive learners.\footnote{
\citet{Gotte2022} and \citet{Ozyilmaz2022} are exceptions that study settings with endogenous information processes.} They 
observe a noisy signal and report their beliefs over subsequent rounds.\footnote{\citet{Bracha2012} and 
\citet{Mobius2022} are some examples}. Although these studies provide important insights, they are 
not flexible enough 
to incorporate the richer theories that have been proposed more recently. In particular, they 
do not allow the studies of endogenous learning or for learning about multiple parameters at once.\footnote{\citet{Coutts2020} 
studies an environment with an ego-relevant parameter and an exogenous state but does not 
incorporate the endogenous information process.}

In my experiment, I move away from the standard framework to study a richer set of 
underlying mechanisms. In particular, 
the interaction between the two parameters, together with an endogenous information process, 
give rise to three possible explanations
for the persistence of incorrect beliefs: dogmatism, the existence of learning traps and misattribution bias.

Within the first explanation, even with perfect Bayesian updating the agent 
may fall into learning traps.\footnote{Conditions under which this happens are detailed in \citet{Hestermann2021}} 
These traps are characterized by a 
combination of misspecified beliefs and 
optimal actions which produce information that confirms the misspecified belief---this is a 
\emph{self-defeating equilibrium}. 
Once an agent falls into it, the belief will be stable and even with 
a correctly specified model of learning, they will not be able to abandon their misspecified beliefs. 
If, in addition the agent is 
dogmatic about their belief, @Heidhues2018 show that they will inevitably fall into a trap
and thus will be able to rationalize and sustein their initial dogmatically overconfident 
(or underconfident) belief.\footnote{\citet{Gotte2022} study the case of agents with dogmatic initial beliefs in a 
laboratory experiment.}

@Ba2023 moves away from dogmatism and endows the agents with a mechanism through which 
they can abandon misspecified beliefs; this allows them to avoid falling into learning traps. 
To do so agents perform Bayesian hypothesis tests to evaluate which is the more 
likely parameter out of two possibilities. By doing this, she characterizes the set of situations 
in which, even agents who consider alternative paradigms, may become trapped.\footnote{A 
similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion.}

Lastly, misattribution bias is the more classical explanation which has been widely studied in 
behavioral science.\footnote{see \citet{kelley1980} for a review.}
Agents who suffer from misattribution bias will attribute successes to their own ability---the 
ego-relevant parameter---and failures to bad luck---the state. 
Under this model of learning, even an agent who initially has a correct initial 
belief may become overconfident if they observe a sequence of successes. In this case, 
the main driver of the bias is not an initial misspecification or the presence of learning traps, 
it is the updating procedure itself.\footnote{A more general framework that can be used to model 
this bias has also been proposed by \citet{Brunnermeier2005} and empirically studied by \citet{Bracha2012}.}

These theories provide the main building blocks for a simplified framework that can be directly 
implemented in a laboratory experiment. The goal is to 
identify the which of the 3 forces---the presence of learning traps, mispecified initial beliefs, or missatribution 
bias---better explains the observed behavior. To determine the fit of the models, I compare the behavior
predicted by each theory to the benchmark given by the fully Bayesian updating procedure.

I also study whether the misspecification is inherently linked to the ego-relevance the parameters 
or a more general phenomenon. I vary the degree of ego-relevance by introducing a 
treatment in which subjects learn about the ability of another participant.
In this treatment, the participants know only the gender and nationality of the other, which can 
induce stereotypes---a different type of misspecification.  
If correct learning about the parameters happens at higher rates in the stereotype treatment, 
it would suggest that the 
bias is intrinsically linked to the ego-relevance of the parameter. In contrast, if similar 
biased behavior arises in both treatments, it is more likely that the main driver of these types 
of misspecified beliefs is the updating procedure itself or the endogenous information process.

Although some agents do fall into learning traps, I find that the behavior of most 
subjects is better explained by misattribution bias. I also find that misattribution is 
no more prevalent in the ego-relevant than in the stereotype condition. This suggests 
that the main driver of the misspecification is the updating proceduce. However, the underlying 
mechanism by which the bias is generated may be 
different in both treatments: While in the ego-relevant condition, the intuitive explanation behind the
bias is that the agent prefers to hold themselves in high esteem, in the stereotype condition
it seems to be driven by some sort of self-censoring---when subjects realize that they underestimated 
the ability of another participant based on their gender and nationality, they over-compensate by
overestimating their ability.

Finally, I estimate the structural parameters of the models to study model-heterogeneity in the sample.
I find that even at an individual level the behavior is better explained by misattribution bias for most
subjects. There is a smaller group of subjects that can be better explained through dogmatic beliefs
and hypothesis testing and none of them behave in line with the fully Bayesian benchmark.

In what follows I first discuss the theoretical framework and the predictions of each of the theories.
Then I introduce a unifying example and my hypotheses. I then describe the experimental
design and the results, as well as the estimation of the parameters.

>>>>

---
title: "Learning with Misspecified Models: the case of overestimation"  
date: '`r format(Sys.time(), "%B %d, %Y")`'
author: "Jimena Galindo"
abstract: "I design a framework and a laboratory experiment that allows for the comparison of multiple theories of misspecified learning. I focus on a framework with endogenous information and a data-generating process ruled by two fundamentals: an ego-relevant parameter and a state. Within this framework I study three forces that can lead to misspecified beliefs: initial misspecifications, learning traps and biased updating. I find that biased updating is the main driver of misspecified beliefs in the lab. In addition, I vary the degree of ego-relevance of the parameter by introducing a stereotype treatment. The data is consistent with biased updating in both cases but for potentially different reasons: when learning about themselves, subjects attribute successes to their own ability and failures to luck. Instead, in the stereotype treatment, they compensate for initial negative biases by over-attributing positive signals to the ability of others. This translates into similar observed choices but different dynamics in beliefs."

#classoption: pagebackref # passes pagebackref=true to hyperref before it is loaded, allowing backreferencing
output: 
    pdf_document:
        citation_package: natbib
        fig_width: 7  
        fig_height: 6
        fig_caption: true
        number_sections: true

@@

A growing body of literature in economics explores how people develop incorrect beliefs 
about fundamentals. Most of this research centers on scenarios where agents 
passively observe the world, and incorrectly integrate the received information into their 
beliefs.\footnote{See \citet{benjamin2019} for a review of the literature on errors of 
probabilistic thinking.}
However, many real-world situations cast agents as active participants in the generation of 
information. In these cases, the information they observe is influenced by their actions and 
subsequent behavior is in turn determined by how they incorporate the information 
into their beliefs.

As an example consider a student who needs to decide how much effort to put into 
studying for an exam. Their decision will depend on two factors: their belief about their intrinsic 
ability, and their belief about how difficult the exam will be. The outcome 
they observe will be affected by how much they decide to study. Imagine that the student 
puts in a moderate amount of effort and gets a surprisingly good grade. Did they get a good grade because 
they are smarter than they thought? Or because the exam was easier than they had anticipated? 
Their future exam preparation strategy will depend fundamentally on which line of reasoning they 
take. This feedback loop is referred to as an \emph{endogenous information 
process} and is at the center of the forces I study.

To understand what the main forces at play are, I compare a set of theories that model learning in settings with 
endogenous information, and which can rationalize the persistence of misspecified beliefs. I develop a unifying framework that nests multiple 
theories of learning and generates testable predictions for each of them. 
Then I cast this framework in a laboratory experiment and test the predictions to identify which 
of the theories are consistent with the behavior I observe in the lab.
The experiment features an agent who needs to 
learn two parameters: one that pertains their own characteristics 
(an \emph{ego-relevant} parameter), 
and an exogenously determined state---in the context of the student preparing for exams, 
the two parameters  would be their intrinsic ability and the difficulty of the exam.

@@

In all of these examples,
holding an incorrect belief about a fundamental leads to suboptimal choices with 
potentially high costs. In spite of the abundance of evidence, the scope of the existing research 
is limited in terms of the frameworks it considers. Most of the experimental evidence 
documenting the bias is collected in settings where subject are passive learners.\footnote{
\citet{Gotte2022} and \citet{Ozyilmaz2022} are exceptions that study settings with endogenous information processes.} They 
observe a noisy signal and report their beliefs over subsequent rounds.\footnote{\citet{Bracha2012} and 
\citet{Mobius2022} are some examples.} Although these studies provide important insights, they are 
not flexible enough 
to incorporate the richer theories that have been proposed more recently. In particular, they 
do not allow the study of endogenous learning or for learning about multiple parameters at once.\footnote{\citet{Coutts2020} 
studies an environment with an ego-relevant parameter and an exogenous state but does not 
incorporate the endogenous information process.}

In my experiment, I move away from the standard framework of passive learning to analyze a richer set of learning mechanisms. In particular, 
the interaction between the two parameters together with an endogenous information process 
gives rise to three possible mechanisms that allow
for the persistence of incorrect beliefs: the presence of learning traps, incorrect initial beliefs, and misattribution bias. The theories that I consider incorporate different combinations of these mechanisms.

When the setting features learning traps, even an agent who incorporates all information correctly, 
may fall into learning traps as outlined by @Hestermann2021. These traps are characterized by a 
combination of an incorrect belief and an
optimal action which produce information that confirms the incorrect belief. 
Once an agent falls into a trap, the belief will be stable and even with 
a correctly specified model of learning, they will not be able to abandon their misspecified beliefs. 
If, the agent is 
dogmatic about their initial belief, @Heidhues2018 show that they will inevitably fall into a trap
and thus will be able to rationalize and sustain their initial misspecification belief.\footnote{\citet{Gotte2022} study the case of agents with dogmatic initial beliefs in a 
laboratory experiment.}

@Ba2023 moves away from dogmatism and endows the agents with a mechanism through which 
they can abandon incorrect beliefs; this allows them to avoid falling into learning traps. 
To do so agents perform Bayesian hypothesis tests that evaluate which is the more 
likely parameter out of two possibilities. By doing this, she characterizes the set of situations 
in which, even agents who consider alternative paradigms, may become trapped.\footnote{A 
similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion.}

Lastly, misattribution bias is the more classical explanation and has been widely studied in 
behavioral science.\footnote{See \citet{kelley1980} for a review.}
Agents who suffer from misattribution bias will attribute successes to their own ability---the 
ego-relevant parameter---and failures to bad luck---the state. 
Under this model of learning, even an agent who initially has a correct initial 
belief may become overconfident if they observe a sequence of successes. In this case, 
the main driver of the bias is not an initial misspecification or the presence of learning traps, 
it is the updating procedure itself.\footnote{A more general framework that can be used to model 
this bias has also been proposed by \citet{Brunnermeier2005} and empirically studied by \citet{Bracha2012}.}

These theories provide the main building blocks for a simplified framework that can be directly 
implemented in a laboratory experiment. In the experiment subjects make choices and receive feedback that depends on their own ability, an exogenous parameter and the choice they made. I track their choices as well as their beliefs about their own ability. The goal is to 
identify which of the 3 forces---the presence of learning traps, misspecified initial beliefs, or missatribution 
bias---better explains the observed behavior. To determine the fit of the models, I compare the behavior
predicted by each theory to the benchmark given by the fully Bayesian updating procedure.

I also study whether the learning mechanism is inherently linked to the ego-relevance the parameters 
or if it is a more general phenomenon. I vary the degree of ego-relevance by introducing a 
treatment in which subjects learn about the ability of another participant.
In this treatment, the participants know only the gender and nationality of the other, and thus can 
induce stereotypes---a different type of misspecification.  
If correct learning about the parameters happens at higher rates in the stereotype treatment, 
it would suggest that the 
bias is intrinsically linked to the ego-relevance of the parameter. In contrast, if similar 
biased behavior arises in both treatments, it is more likely that the main driver of these types 
of misspecified beliefs is the updating procedure itself or the endogenous information process.

Although some agents do fall into learning traps, I find that the behavior of most 
subjects is better explained by misattribution bias: good news are treated as signaling high ability, while bad news are attributed to a low state. I also find that misattribution is 
no more prevalent in the ego-relevant than in the stereotype condition. This suggests 
that the main driver of the misspecification is the updating proceduce; however, the underlying 
mechanism by which the bias is generated may be 
different in both treatments: While in the ego-relevant condition subjects prefer to hold themselves in high esteem, in the stereotype condition
updating seems to be driven by some sort of bias overcorrection---when subjects realize that they underestimated 
the ability of another participant based on their gender and nationality, they compensate by
overestimating their ability.

Finally, I estimate the structural parameters of the models to study model-heterogeneity in the sample.
I find that even at an individual level the behavior is better explained by a general model of 
misattribution bias for most
subjects. There is a smaller group of subjects that can be better explained through dogmatic beliefs
and hypothesis testing and none of them behave in line with the fully Bayesian benchmark.

In what follows I first discuss the theoretical framework and the predictions of each of the theories.
Then I introduce a unifying example and my hypotheses. In section 4 I describe the experimental
design and in section 5 I present the data and the results. Section 6 outlines the estimation of the parameters
and the model fit analysis.

### 2023-10-14 Paper/draft.tex
Edit intro and abstract

\newcommand{\EE}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\PP}[1]{\mathbb{P}\left(#1\right)}
\newcommand{\CP}[2]{\mathbb{P}\left(#1 \,| \, #2 \right)}
\newcommand{\CE}[2]{\mathbb{E}\left[#1\,|\,#2 \right]}
\usepackage[bottom]{footmisc}
\usepackage[doublespacing]{setspace}
\usepackage[normal]{caption}
\usepackage{dsfont}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{hyperref}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Learning with misspecified models: the case of overestimation},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Learning with misspecified models: the case of overestimation}
\author{Jimena Galindo}
\date{October 14, 2023}

\begin{document}
\maketitle
\begin{abstract}
I design a framework and laboratory experiment that allows for the
comparison of multiple theories of misspecified learning. I focus on a
framework with endogenous information and two a data generating process
ruled by two fundamentals: an ego-relevant parameter and a state. Within
this framework I study three forces that can lead to misspecified
beliefs: initial misspecifications, learning traps and biased updating.
I find that biased updating is the main driver of misspecifications. In
addition, I vary the degree of ego-relevance of the parameter by
introducing a stereotype treatment. The data is consistent with biased
updating in both cases but for potentially different reasons: In the
ego-relevant treatment subjects attribute successes to their own ability
and failures to luck; in the stereotype treatment they over-compensate
for initial negative biases by over-attributing positive signals to the
ability of others. This translates into similar observed choices but
different dynamic in beliefs.
\end{abstract}

@@

In all of these examples, holding an incorrect belief about a
fundamental leads to suboptimal choices with potentially high costs. In
spite of the abundance of evidence, the scope of the existing research
is limited in terms of the frameworks it considers. Most of the
experimental evidence documenting the bias is collected in settings
where subject are passive learners.\footnote{
\citet{Gotte2022} and \citet{Ozyilmaz2022} are exceptions that study settings with endogenous information processes.}
They observe a noisy signal and report their beliefs over subsequent
rounds.\footnote{\citet{Bracha2012} and 
\citet{Mobius2022} are some examples}. Although these studies provide
important insights, they are not flexible enough to incorporate the
richer theories that have been proposed more recently. In particular,
they do not allow the studies of endogenous learning or for learning
about multiple parameters at once.\footnote{\citet{Coutts2020} 
studies an environment with an ego-relevant parameter and an exogenous state but does not 
incorporate the endogenous information process.}

In my experiment, I move away from the standard framework to study a
richer set of underlying mechanisms. In particular, the interaction
between the two parameters, together with an endogenous information
process, give rise to three possible explanations for the persistence of
incorrect beliefs: dogmatism, the existence of learning traps and
misattribution bias.

Within the first explanation, even with perfect Bayesian updating the
agent may fall into learning
traps.\footnote{Conditions under which this happens are detailed in \citet{Hestermann2021}}
These traps are characterized by a combination of misspecified beliefs
and optimal actions which produce information that confirms the
misspecified belief---this is a \emph{self-defeating equilibrium}. Once
an agent falls into it, the belief will be stable and even with a
correctly specified model of learning, they will not be able to abandon
their misspecified beliefs. If, in addition the agent is dogmatic about
their belief, \citet{Heidhues2018} show that they will inevitably fall
into a trap and thus will be able to rationalize and sustein their
initial dogmatically overconfident (or underconfident)
belief.\footnote{\citet{Gotte2022} study the case of agents with dogmatic initial beliefs in a 
laboratory experiment.}

\citet{Ba2023} moves away from dogmatism and endows the agents with a
mechanism through which they can abandon misspecified beliefs; this
allows them to avoid falling into learning traps. To do so agents
perform Bayesian hypothesis tests to evaluate which is the more likely
parameter out of two possibilities. By doing this, she characterizes the
set of situations in which, even agents who consider alternative
paradigms, may become trapped.\footnote{A 
similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion.}

Lastly, misattribution bias is the more classical explanation which has
been widely studied in behavioral
science.\footnote{see \citet{kelley1980} for a review.} Agents who
suffer from misattribution bias will attribute successes to their own
ability---the ego-relevant parameter---and failures to bad luck---the
state. Under this model of learning, even an agent who initially has a
correct initial belief may become overconfident if they observe a
sequence of successes. In this case, the main driver of the bias is not
an initial misspecification or the presence of learning traps, it is the
updating procedure
itself.\footnote{A more general framework that can be used to model 
this bias has also been proposed by \citet{Brunnermeier2005} and empirically studied by \citet{Bracha2012}.}

These theories provide the main building blocks for a simplified
framework that can be directly implemented in a laboratory experiment.
The goal is to identify the which of the 3 forces---the presence of
learning traps, mispecified initial beliefs, or missatribution
bias---better explains the observed behavior. To determine the fit of
the models, I compare the behavior predicted by each theory to the
benchmark given by the fully Bayesian updating procedure.

I also study whether the misspecification is inherently linked to the
ego-relevance the parameters or a more general phenomenon. I vary the
degree of ego-relevance by introducing a treatment in which subjects
learn about the ability of another participant. In this treatment, the
participants know only the gender and nationality of the other, which
can induce stereotypes---a different type of misspecification.\\
If correct learning about the parameters happens at higher rates in the
stereotype treatment, it would suggest that the bias is intrinsically
linked to the ego-relevance of the parameter. In contrast, if similar
biased behavior arises in both treatments, it is more likely that the
main driver of these types of misspecified beliefs is the updating
procedure itself or the endogenous information process.

Although some agents do fall into learning traps, I find that the
behavior of most subjects is better explained by misattribution bias. I
also find that misattribution is no more prevalent in the ego-relevant
than in the stereotype condition. This suggests that the main driver of
the misspecification is the updating proceduce. However, the underlying
mechanism by which the bias is generated may be different in both
treatments: While in the ego-relevant condition, the intuitive
explanation behind the bias is that the agent prefers to hold themselves
in high esteem, in the stereotype condition it seems to be driven by
some sort of self-censoring---when subjects realize that they
underestimated the ability of another participant based on their gender
and nationality, they over-compensate by overestimating their ability.

Finally, I estimate the structural parameters of the models to study
model-heterogeneity in the sample. I find that even at an individual
level the behavior is better explained by misattribution bias for most
subjects. There is a smaller group of subjects that can be better
explained through dogmatic beliefs and hypothesis testing and none of
them behave in line with the fully Bayesian benchmark.

In what follows I first discuss the theoretical framework and the
predictions of each of the theories. Then I introduce a unifying example
and my hypotheses. I then describe the experimental design and the
results, as well as the estimation of the parameters.

>>>>

\newcommand{\EE}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\PP}[1]{\mathbb{P}\left(#1\right)}
\newcommand{\CP}[2]{\mathbb{P}\left(#1 \,| \, #2 \right)}
\newcommand{\CE}[2]{\mathbb{E}\left[#1\,|\,#2 \right]}
\usepackage[bottom]{footmisc}
\usepackage[doublespacing]{setspace}
\usepackage[normal]{caption}
\usepackage{dsfont}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{hyperref}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Learning with Misspecified Models: the case of overestimation},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Learning with Misspecified Models: the case of overestimation}
\author{Jimena Galindo}
\date{October 14, 2023}

\begin{document}
\maketitle
\begin{abstract}
I design a framework and a laboratory experiment that allows for the
comparison of multiple theories of misspecified learning. I focus on a
framework with endogenous information and a data-generating process
ruled by two fundamentals: an ego-relevant parameter and a state. Within
this framework I study three forces that can lead to misspecified
beliefs: initial misspecifications, learning traps and biased updating.
I find that biased updating is the main driver of misspecified beliefs
in the lab. In addition, I vary the degree of ego-relevance of the
parameter by introducing a stereotype treatment. The data is consistent
with biased updating in both cases but for potentially different
reasons: when learning about themselves, subjects attribute successes to
their own ability and failures to luck. Instead, in the stereotype
treatment, they compensate for initial negative biases by
over-attributing positive signals to the ability of others. This
translates into similar observed choices but different dynamics in
beliefs.
\end{abstract}

@@

In all of these examples, holding an incorrect belief about a
fundamental leads to suboptimal choices with potentially high costs. In
spite of the abundance of evidence, the scope of the existing research
is limited in terms of the frameworks it considers. Most of the
experimental evidence documenting the bias is collected in settings
where subject are passive learners.\footnote{
\citet{Gotte2022} and \citet{Ozyilmaz2022} are exceptions that study settings with endogenous information processes.}
They observe a noisy signal and report their beliefs over subsequent
rounds.\footnote{\citet{Bracha2012} and 
\citet{Mobius2022} are some examples.} Although these studies provide
important insights, they are not flexible enough to incorporate the
richer theories that have been proposed more recently. In particular,
they do not allow the study of endogenous learning or for learning about
multiple parameters at once.\footnote{\citet{Coutts2020} 
studies an environment with an ego-relevant parameter and an exogenous state but does not 
incorporate the endogenous information process.}

In my experiment, I move away from the standard framework of passive
learning to analyze a richer set of learning mechanisms. In particular,
the interaction between the two parameters together with an endogenous
information process gives rise to three possible mechanisms that allow
for the persistence of incorrect beliefs: the presence of learning
traps, incorrect initial beliefs, and misattribution bias. The theories
that I consider incorporate different combinations of these mechanisms.

When the setting features learning traps, even an agent who incorporates
all information correctly, may fall into learning traps as outlined by
\citet{Hestermann2021}. These traps are characterized by a combination
of an incorrect belief and an optimal action which produce information
that confirms the incorrect belief. Once an agent falls into a trap, the
belief will be stable and even with a correctly specified model of
learning, they will not be able to abandon their misspecified beliefs.
If, the agent is dogmatic about their initial belief,
\citet{Heidhues2018} show that they will inevitably fall into a trap and
thus will be able to rationalize and sustain their initial
misspecification
belief.\footnote{\citet{Gotte2022} study the case of agents with dogmatic initial beliefs in a 
laboratory experiment.}

\citet{Ba2023} moves away from dogmatism and endows the agents with a
mechanism through which they can abandon incorrect beliefs; this allows
them to avoid falling into learning traps. To do so agents perform
Bayesian hypothesis tests that evaluate which is the more likely
parameter out of two possibilities. By doing this, she characterizes the
set of situations in which, even agents who consider alternative
paradigms, may become trapped.\footnote{A 
similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion.}

Lastly, misattribution bias is the more classical explanation and has
been widely studied in behavioral
science.\footnote{See \citet{kelley1980} for a review.} Agents who
suffer from misattribution bias will attribute successes to their own
ability---the ego-relevant parameter---and failures to bad luck---the
state. Under this model of learning, even an agent who initially has a
correct initial belief may become overconfident if they observe a
sequence of successes. In this case, the main driver of the bias is not
an initial misspecification or the presence of learning traps, it is the
updating procedure
itself.\footnote{A more general framework that can be used to model 
this bias has also been proposed by \citet{Brunnermeier2005} and empirically studied by \citet{Bracha2012}.}

These theories provide the main building blocks for a simplified
framework that can be directly implemented in a laboratory experiment.
In the experiment subjects make choices and receive feedback that
depends on their own ability, an exogenous parameter and the choice they
made. I track their choices as well as their beliefs about their own
ability. The goal is to identify which of the 3 forces---the presence of
learning traps, misspecified initial beliefs, or missatribution
bias---better explains the observed behavior. To determine the fit of
the models, I compare the behavior predicted by each theory to the
benchmark given by the fully Bayesian updating procedure.

I also study whether the learning mechanism is inherently linked to the
ego-relevance the parameters or if it is a more general phenomenon. I
vary the degree of ego-relevance by introducing a treatment in which
subjects learn about the ability of another participant. In this
treatment, the participants know only the gender and nationality of the
other, and thus can induce stereotypes---a different type of
misspecification.\\
If correct learning about the parameters happens at higher rates in the
stereotype treatment, it would suggest that the bias is intrinsically
linked to the ego-relevance of the parameter. In contrast, if similar
biased behavior arises in both treatments, it is more likely that the
main driver of these types of misspecified beliefs is the updating
procedure itself or the endogenous information process.

Although some agents do fall into learning traps, I find that the
behavior of most subjects is better explained by misattribution bias:
good news are treated as signaling high ability, while bad news are
attributed to a low state. I also find that misattribution is no more
prevalent in the ego-relevant than in the stereotype condition. This
suggests that the main driver of the misspecification is the updating
proceduce; however, the underlying mechanism by which the bias is
generated may be different in both treatments: While in the ego-relevant
condition subjects prefer to hold themselves in high esteem, in the
stereotype condition updating seems to be driven by some sort of bias
overcorrection---when subjects realize that they underestimated the
ability of another participant based on their gender and nationality,
they compensate by overestimating their ability.

Finally, I estimate the structural parameters of the models to study
model-heterogeneity in the sample. I find that even at an individual
level the behavior is better explained by a general model of
misattribution bias for most subjects. There is a smaller group of
subjects that can be better explained through dogmatic beliefs and
hypothesis testing and none of them behave in line with the fully
Bayesian benchmark.

In what follows I first discuss the theoretical framework and the
predictions of each of the theories. Then I introduce a unifying example
and my hypotheses. In section 4 I describe the experimental design and
in section 5 I present the data and the results. Section 6 outlines the
estimation of the parameters and the model fit analysis.
